OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-06-25 paper/paper.Rmd
Minor edits

Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
important step in characterizing the composition of microbial communities across
large datasets.
OptiClust, a _de novo_ OTU clustering method in the mothur program, has been
shown to produce higher quality OTU assignments than other methods and at
comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent methods
is that OTU assignments clustered with _de novo_ methods may change when new
sequences are added to a dataset.
However, in some cases one may wish to incorporate new samples into a previously
clustered dataset without performing clustering again on all sequences, such as
when comparing across datasets or deploying machine learning models where OTUs
are features.
To provide an efficient and robust method to fit amplicon sequence data to
existing OTUs, we developed the OptiFit algorithm.
We tested OptiFit using four microbiome datasets with two different strategies:
by fitting to an external reference database, or by splitting the dataset into a 
reference and query set and fitting the query sequences to the reference set.
The result is a robust implementation of closed and open-refernce clustering.
OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
using the split dataset strategy, although the OTU quality and processing speed
depends on the database chosen when using the external database strategy.
OptiFit provides a suitable option for users who require consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

**TODO.**
**"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**

\newpage

@@

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To test the best reference size, reference sets with 10% to 80% of the
sequences were created, with the remaining sequences used for the query.
OTU quality was remarkably stable across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all
fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` with 10% of
sequences in the reference and `r sec_fit_split_human_simple_9` with 90% of
sequences in the reference (Figure \ref{fig:results_split}).
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and 
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference. 
These trends held for the other datasets as well (Figure 
\ref{fig:results_split}).
Thus, the reference fraction doid not affect OTU quality in terms of MCC score,
but did affect the run time and the fraction of sequences that mapped during the
closed reference clustering.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which contains only 21,000 (**TODO insert exact number**)
sequences compared to over 200,000 (**TODO insert exact number**) sequences in SILVA and Greengenes each.
Note that Greengenes has not been updated since 2013 at the time of this
writing, while SILVA is updated every one to two years with the most recent
release in 2019.
We recommend that users who require an independent reference database opt for
large databases with regular updates and good coverage of microbial diversity.
Since OptiClust still performs faster than open reference OptiFit and creates
higher quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever stable OTUs
are not required for the study at hand.

@@

When fitting with the split dataset strategy, OTU quality was remarkably similar
when reference sequences were selected by a simple random sample or weighted by
abundance, but quality was slightly worse when sequences were weighted by
similarity. 
We recommend using a simple random sample since the more
sophisticated reference selection methods do not offer any benefit. 
The similarity in OTU quality between OptiClust and OptiFit with this strategy
demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
such as when comparing OTUs across studies.
However, when stable OTUs are not required, we recommend using OptiClust for _de
novo_ clustering over the split strategy with OptiFit since OptiClust is simpler
to execute but performs similarly in terms of both run time and OTU quality.

We have developed a new clustering algorithm that allows users to produce high
quality OTUs using already existing OTUs as a reference.
Unlike existing reference-based methods that map query sequences to a single
centroid sequence in each reference OTU, OptiFit considers all sequences in each
reference OTU when fitting query sequences, resulting in OTUs of a similar high
quality as those produced by the _de novo_ OptiClust algorithm.
Potential applications include fitting sequences to reference databases,
comparing taxonomic composition of microbiomes across different studies, 
or using OTU-based machine learning models to make predictions on new data.
OptiFit fills the missing option for fitting query sequences to existing OTUs
that does not sacrifice OTU quality for stability of OTU assignments.

@@

We downloaded 16S rRNA gene amplicon sequences from four published datasets
isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
These were selected because **TODO**.
Raw sequences were processed using mothur according to the Schloss Lab MiSeq SOP
as described in the mothur wiki and accompanying study by Kozich _et al._
[@schloss_miseq_nodate; @kozich_development_2013].
These steps included trimming and filtering for quality, aligning to the SILVA
reference alignment [@quast_silva_2013], discarding sequences that aligned
outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
and calculating distances between all pairs of sequences within each dataset
prior to clustering.

>>>>

Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
important step in characterizing the composition of microbial communities across
large datasets.
OptiClust, a _de novo_ OTU clustering method in the mothur program, has been
shown to produce higher quality OTU assignments than other methods and at
comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent methods
is that OTU assignments clustered with _de novo_ methods may change when new
sequences are added to a dataset.
However, in some cases one may wish to incorporate new samples into a previously
clustered dataset without performing clustering again on all sequences, such as
when comparing across datasets or deploying machine learning models where OTUs
are features.
To provide an efficient and robust method to fit amplicon sequence data to
existing OTUs, we developed the OptiFit algorithm.
We tested OptiFit using four microbiome datasets with two different strategies:
by fitting to an external reference database, or by splitting the dataset into a 
reference and query set and fitting the query sequences to the reference set.
The result is a robust implementation of closed and open-reference clustering.
OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
using the split dataset strategy, although the OTU quality and processing speed
depends on the database chosen when using the external database strategy.
OptiFit provides a suitable option for users who require consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

**TODO.**
<!--
**"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
Advancements in sequencing technology have allowed researchers to affordably generate millions of reads of microorganisms from diverse natural communities,
allowing us to characterize communities.
-->
\newpage

@@

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To test the best reference size, reference sets with 10% to 80% of the
sequences were created, with the remaining sequences used for the query.
OTU quality was remarkably consistent across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all
fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` with 10% of
sequences in the reference and `r sec_fit_split_human_simple_9` with 90% of
sequences in the reference (Figure \ref{fig:results_split}).
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and 
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference. 
These trends held for the other datasets as well (Figure 
\ref{fig:results_split}).
Thus, the reference fraction doid not affect OTU quality in terms of MCC score,
but did affect the run time and the fraction of sequences that mapped during the
closed reference clustering.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which contains only 21,000 (**TODO insert exact number**)
sequences compared to over 200,000 (**TODO insert exact number**) sequences in SILVA and Greengenes each.
Note that Greengenes has not been updated since 2013 at the time of this
writing, while SILVA is updated every one to two years with the most recent
release in 2019.
We recommend that users who require an independent reference database opt for
large databases with regular updates and good coverage of microbial diversity.
Since OptiClust still performs faster than open reference OptiFit and creates
higher quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever consistent
OTUs are not required for the study at hand.

@@

When fitting with the split dataset strategy, OTU quality was remarkably similar
when reference sequences were selected by a simple random sample or weighted by
abundance, but quality was slightly worse when sequences were weighted by
similarity. 
We recommend using a simple random sample since the more
sophisticated reference selection methods do not offer any benefit. 
The similarity in OTU quality between OptiClust and OptiFit with this strategy
demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
such as when comparing OTUs across studies.
However, when consistent OTUs are not required, we recommend using OptiClust for
_de novo_ clustering over the split strategy with OptiFit since OptiClust is
simpler to execute but performs similarly in terms of both run time and OTU
quality.

We have developed a new clustering algorithm that allows users to produce high
quality OTUs using already existing OTUs as a reference.
Unlike existing reference-based methods that map query sequences to a single
centroid sequence in each reference OTU, OptiFit considers all sequences in each
reference OTU when fitting query sequences, resulting in OTUs of a similar high
quality as those produced by the _de novo_ OptiClust algorithm.
Potential applications include fitting sequences to reference databases,
comparing taxonomic composition of microbiomes across different studies, 
or using OTU-based machine learning models to make predictions on new data.
OptiFit fills the missing option for fitting query sequences to existing OTUs
that does not sacrifice OTU quality for consistency of OTU assignments.

@@

We downloaded 16S rRNA gene amplicon sequences from four published datasets
isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
These datasets represent a selection of the broad types of natural communities
that microbial ecologists study.
We processed the raw sequences using mothur according to the Schloss Lab MiSeq
SOP as described in the mothur wiki and accompanying study by Kozich _et al._
[@schloss_miseq_nodate; @kozich_development_2013].
These steps included trimming and filtering for quality, aligning to the SILVA
reference alignment [@quast_silva_2013], discarding sequences that aligned
outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
and calculating distances between all pairs of sequences within each dataset
prior to clustering.
